A Type System for Safe and Scalable Parallelism using Logical Regions \\ (Extended Version)

### 2012-07-19 doc/popl2013full/langdef.tex
unpack, a few more typo fixes

The proof is a standard induction on the structure of the derivation
For each of 
the Core Legion expressions there are six separate parts to the proof,
one for each of the three consistency properties that must be shown
for the execution environment of subexpressions and one for each of
the three properties of the execution's results.
Many of the cases are similar, and benefit from
the use of the following straightforward lemmas:
\begin{lem}
\label{lemma:heapconst:apply}
\rm
If $\storeconsist{S}{H}$ and $\traceconsist{E}$, then $\storeconsist{\text{apply}(S, E)}{H}$.
\end{lem}

@@

\vspace{4pt}
\hangindent=2\parindent \hangafter=1 \noindent %
$\storeconsist{S}{H}$ -  The heap typing consistency of all stores used in subexpressions follows directly from
Lemma~\ref{lem:heapconst:apply}.

@@

\vspace{4pt}
\hangindent=2\parindent \hangafter=1 \noindent %
$\traceconsist{E}$ -  The type consistency of the values in an expression's memory
trace follows from Lemma~\ref{lem:heapconst:effects1} and Lemma~\ref{lem:heapconst:effects2}.
New memory operations are added by {\tt write} and {\tt reduce} expressions, but consistency follows
directly from the induction hypothesis.  Finally, the consistency of the
values in a called function's memory trace is addressed in the same way as the return value.

@@

\item Case [E-Newcolor].
By assumption, 
$$\opsenv{\text{newcolor } r}{K}{[]}$$ 
where  $K = \{ (l_1,v_1),\ldots,(l_p,i_p) \}$ such that
$l_i \in M(r)$ for all $i$ and $l_1,\ldots,l_p$ are pairwise distinct locations.
In addition, we have the assumptions
$$\typeenv{\text{newcolor } r}{T}$$ and
$\mapconsist{\Omega}$ and 
$\localconsist{L}{\Gamma}$ and 
$\storeconsist{S}{H}$.
The only type rule that can apply is [T-Newcolor], so we can infer that $T = \text{coloring}(r)$.
From the structure of $K$ it follows that $\valueconsist{K}{\text{coloring(r)}}$.
The other two conclusions,
$\traceconsist{[]}$ and 
$\privconsist{[]}{\Phi}$, follow trivially.

@@

To show $\valueconsist{v}{T'@(\oton{r})}$, we first observe that from
$\valueconsist{v}{T'@(r_1',\ldots,r_k')}$ we know
there is an $i$ such that $v \in M(r_i')$ and 
$H(v) = \hat T'$.   From [T-UpRgn]
we know $\forall i. \exists j. r_i' \leq r_j \in \Omega^*$.  Combining this with
$\mapconsist{\Omega}$ and 
From Lemma~\ref{lemma:omegaclosure}, we have $mapconsist{\Omega^*}$, so there
must be some $r_j$ such that $M(r_i') \subseteq M(r_j)$.  Therefore $v \in M(r_j)$
and $\valueconsist{v}{T'@(\oton{r})}$.

@@

Similarly, observe that
$$\hat \Omega_1 [ \rho_1/r'_1,\ldots,\rho_k/r'_k ] = M \llbracket \Omega_1 [ r_1/r'_1,\ldots,r_k/r'_k ] \rrbracket$$
.  With two applications of Lemma~\ref{lemma:mappedomega}, we have $\mapconsist[\{(r'_i,\rho_i)\}]{\hat \Omega_1}$
if and only if $\mapconsist{\Omega_1 [ r_1/r'_1,\ldots,r_k/r'_k ]}$.  The latter constraints are a subset of
$\Omega^*$, and must be satisfied if $\mapconsist{\Omega^*}$, which we have from Lemma~\ref{lemma:omegaclosure}.

@@

For subexpression $e_1$, from [E-UnPack] we have $$\opsenv{e_1}{\langle \langle \otok{k}{\rho},v \rangle\rangle}{E}$$ and
from [T-UnPack] we know $$\typeenv{e_1}{T_1 = \exists r'_1,\ldots,r'_n.T_2 \text{ where }\Omega_1}$$
By induction, we can conclude that $\valueconsist{\langle \langle \otok{k}{\rho},v \rangle\rangle}{T_1}$ and
$\traceconsist{E}$ and $\privconsist{E}{\Phi}$.

@@

To show $\mapconsist[M']{\Omega'}$, consider a constraint $r \leq r' \in \Omega'$.  There are two cases. If $r \leq r' \in \Omega_1[r_1/r'_1,\ldots,r_k/r'_k]$
then $M'(r) \subseteq M'(r')$ follows from $\valueconsist{\langle \langle \otok{k}{\rho},v \rangle\rangle}{T_1}$.
If $r \leq r' \not\in \Omega_1[r_1/r'_1,\ldots,r_k/r'_k]$ then $r \leq r' \in \Omega$ and $M'(r) = M(r) \subseteq M(r') = M'(r')$.
The reasoning for constraints $r * r'$ is the same.

To show $\localconsist[M']{L'}{\Gamma'}$, we must show
$\valueconsist[M']{L'(x)}{\Gamma'(x)}$ for all identifiers $x$ in
the domain of $\Gamma'$.  For any $id' \neq id$, the statement is equivalent
to showing $\valueconsist[M']{L(x)}{\Gamma(x)}$, which must hold because $M'$ does 
does not change any existing binding in $M$.  For $id$, we have
$\valueconsist[M']{v_1}{T_1[r_1/r'_1,\ldots,r_k/r'_k]}$, which is implied by
$\valueconsist{\langle \langle \otok{k}{\rho},v \rangle\rangle}{T_1}$.

>>>>

The proof is a standard induction on the structure of the derivation.
For each of 
the Core Legion expressions there are six separate parts to the proof,
one for each of the three consistency properties that must be shown
for the execution environment of subexpressions and one for each of
the three properties of the execution's results.
Many of the cases are similar, and benefit from
the use of the following straightforward lemmas:
\begin{lem}
\label{lemma:heapconst:apply}
\rm
If $\storeconsist{S}{H}$ and $\traceconsist{E}$, then $\storeconsist{\text{apply}(S, E)}{H}$.
\end{lem}

@@

\vspace{4pt}
\hangindent=2\parindent \hangafter=1 \noindent %
$\storeconsist{S}{H}$ -  The heap typing consistency of all stores used in subexpressions follows directly from
Lemma~\ref{lemma:heapconst:apply}.

@@

\vspace{4pt}
\hangindent=2\parindent \hangafter=1 \noindent %
$\traceconsist{E}$ -  The type consistency of the values in an expression's memory
trace follows from Lemma~\ref{lemma:heapconst:effects1} and Lemma~\ref{lemma:heapconst:effects2}.
New memory operations are added by {\tt write} and {\tt reduce} expressions, but consistency follows
directly from the induction hypothesis.  Finally, the consistency of the
values in a called function's memory trace is addressed in the same way as the return value.

@@

\item Case [E-Newcolor].
By assumption, 
$$\opsenv{\text{newcolor } r}{K}{[]}$$ 
where  $K = \{ (l_1,iv_1),\ldots,(l_p,iv_p) \}$ such that
$l_i \in M(r)$ for all $i$ and $l_1,\ldots,l_p$ are pairwise distinct locations.
In addition, we have the assumptions
$$\typeenv{\text{newcolor } r}{T}$$ and
$\mapconsist{\Omega}$ and 
$\localconsist{L}{\Gamma}$ and 
$\storeconsist{S}{H}$.
The only type rule that can apply is [T-Newcolor], so we can infer that $T = \text{coloring}(r)$.
From the structure of $K$ it follows that $\valueconsist{K}{\text{coloring(r)}}$.
The other two conclusions,
$\traceconsist{[]}$ and 
$\privconsist{[]}{\Phi}$, follow trivially.

@@

To show $\valueconsist{v}{T'@(\oton{r})}$, we first observe that from
$\valueconsist{v}{T'@(r_1',\ldots,r_k')}$ we know
there is an $i$ such that $v \in M(r_i')$ and 
$H(v) = \hat T'$.   From [T-UpRgn]
we know $\forall i. \exists j. r_i' \leq r_j \in \Omega^*$.
%  Combining this with $\mapconsist{\Omega}$ and 
From Lemma~\ref{lemma:omegaclosure}, we have $\mapconsist{\Omega^*}$, so there
must be some $r_j$ such that $M(r_i') \subseteq M(r_j)$.  Then $v \in M(r_j)$
and $\valueconsist{v}{T'@(\oton{r})}$.

@@

Similarly, observe that
$$\hat \Omega_1 [ \rho_1/r'_1,\ldots,\rho_k/r'_k ] = M \llbracket \Omega_1 [ r_1/r'_1,\ldots,r_k/r'_k ] \rrbracket$$
.  With two applications of Lemma~\ref{lemma:mappedomega}, we have
$$\mapconsist[\{(r'_i,\rho_i)\}]{\hat \Omega_1} \Leftrightarrow \mapconsist{\Omega_1 [ r_1/r'_1,\ldots,r_k/r'_k ]}$$
.  The latter constraints are a subset of
$\Omega^*$, and must be satisfied if $\mapconsist{\Omega^*}$, which we have from Lemma~\ref{lemma:omegaclosure}.

@@

For subexpression $e_1$, from [E-UnPack] we have $$\opsenv{e_1}{\langle \langle \otok{k}{\rho},v_1 \rangle\rangle}{E}$$ and
from [T-Unpack] we know $$\typeenv{e_1}{T_1 = \exists r'_1,\ldots,r'_n.T_2 \text{ where }\Omega_1}$$
By induction, we can conclude that $\valueconsist{\langle \langle \otok{k}{\rho},v \rangle\rangle}{T_1}$ and
$\traceconsist{E}$ and $\privconsist{E}{\Phi}$.

@@

We can show $\mapconsist[M']{\Omega'}$ by showing both $\mapconsist[M']{\Omega}$ and $\mapconsist[M']{\Omega_1[r_1/r'_1,\ldots,r_k/r'_k]}$.  The former follows directly from $\mapconsist{\Omega}$ as $M'$ does not change any existing
mappings in $M$.  The latter requires an argument similar to the [E-Pack] case.  We let
$M \llbracket T_1 \rrbracket = \exists r'_1, \ldots, r'_n.\ \hat T_2\text{ where }\hat \Omega_1$, and conclude that
$\mapconsist[\{ \rho_1/r'_1,\ldots,\rho_k/r'_k \}]{\hat \Omega_1}$.  From Lemma~\ref{lemma:mappedomega} we have 
$\mapconsist[\emptyset]{\hat \Omega_1[\rho_1/r'_1,\ldots,\rho_k/r'_k]}$.  Finally, we observe that 
$$M' \llbracket \Omega_1[r_1/r'_1,\ldots,r_k/r'_k] \rrbracket = \hat \Omega_1[\rho_1/r'_1,\ldots,\rho_k/r'_k]$$
and conclude $\mapconsist[M']{\Omega'}$ with another application of Lemma~\ref{lemma:mappedomega}.

\begin{comment}
To show $\mapconsist[M']{\Omega'}$, consider a constraint $r \leq r' \in \Omega'$.  There are two cases. If $r \leq r' \in \Omega_1[r_1/r'_1,\ldots,r_k/r'_k]$
then $M'(r) \subseteq M'(r')$ follows from $\valueconsist{\langle \langle \otok{k}{\rho},v \rangle\rangle}{T_1}$.
If $r \leq r' \not\in \Omega_1[r_1/r'_1,\ldots,r_k/r'_k]$ then $r \leq r' \in \Omega$ and $M'(r) = M(r) \subseteq M(r') = M'(r')$.
The reasoning for constraints $r * r'$ is the same.
\end{comment}

To show $\localconsist[M']{L'}{\Gamma'}$, we must show
$\valueconsist[M']{L'(x)}{\Gamma'(x)}$ for all identifiers $x$ in
the domain of $\Gamma'$.  For any $id' \neq id$, the statement is equivalent
to showing $\valueconsist[M']{L(x)}{\Gamma(x)}$, which must hold because $M'$ does 
does not change any existing binding in $M$.
For $id$, we have $\valueconsist[]{v_1}{\hat T_2[\rho_1/r'_1,\ldots,\rho_k/r'_k]}$
and again observe that since $r'_1,\ldots,r'_k$ do not appear in $M$, the right hand side
is equal to $M' \llbracket T_2[r_1/r'_1,\ldots,r_k/r'_k] \rrbracket$.

\begin{comment}
$\valueconsist[M']{v_1}{T_1[r_1/r'_1,\ldots,r_k/r'_k]}$, which is implied by
$\valueconsist{\langle \langle \otok{k}{\rho},v \rangle\rangle}{T_1}$.
\end{comment}
